---
title: "BAN 502 Phase 2: Exploratory Analysis of Super Soaker Failures"
author: "Brendan Hudson"
date: "2026-02-16"
output: word_document
---

# Phase 1: Exploratory Data Analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction

This project explores the product testing dataset to look for patterns related to product failure.

No predictive models are used in this phase. The goal is simply to understand the data and see which variables might matter later.

## Load Data

```{r}
train <- read_csv("train.csv")

train$failure <- as_factor(train$failure)
train$product_code <- as_factor(train$product_code)
train$attribute_0 <- as_factor(train$attribute_0)
train$attribute_1 <- as_factor(train$attribute_1)
```

Categorical variables were converted to factors for proper summaries and plots.

## Dataset Overview

```{r}
str(train)
summary(train)
```

This shows the number of observations, variable types, and basic statistics.

## Failure Distribution

```{r}
table(train$failure)
prop.table(table(train$failure))
```

```{r}
ggplot(train, aes(x = failure)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Product Failures",
    x = "Failure Outcome",
    y = "Count"
  )
```

About **21% of the products failed**, so failures are less common than successes.

## Missing Data

```{r}
missing_counts <- colSums(is.na(train))
sort(colSums(is.na(train)), decreasing = TRUE)
missing_counts <- colSums(is.na(train))
barplot(missing_counts,
        main="Missing Values per Variable",
        ylab="Number Missing")
```

Some measurement variables have missing values. This will need attention in Phase 2.

## Numeric Variables vs Failure

There are many measurement variables, so summary statistics were checked first.

```{r}
measurement_vars <- train %>%
  select(starts_with("measurement"))
summary(measurement_vars)
```

A few measurements were then selected to compare distributions.

```{r}
ggplot(train, aes(x = failure, y = loading)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Loading by Failure Status")

ggplot(train, aes(x = failure, y = measurement_5)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Measurement 5 by Failure Status")

ggplot(train, aes(x = failure, y = measurement_10)) +
  geom_boxplot(fill = "lightyellow") +
  labs(title = "Measurement 10 by Failure Status")

ggplot(train, aes(x = failure, y = measurement_17)) +
  geom_boxplot(fill = "lightcoral") +
  labs(title = "Measurement 17 by Failure Status")
```

These plots show visible differences between failed and non-failed products, suggesting some numeric variables may relate to failure.

## Categorical Variables vs Failure

```{r}
table(train$failure, train$product_code)
prop.table(table(train$failure, train$product_code), 2)
```

```{r}
ggplot(train, aes(x = product_code, fill = failure)) +
  geom_bar(position = "fill") +
  labs(
    title = "Failure Rate by Product Type",
    y = "Proportion"
  )
```

Failure rates differ by product type, which may indicate design differences matter.

## Key Findings

- About **21%** of products failed
- Loading and several measurements show different distributions by outcome
- Failure varies across product types
- Missing data is present in several variables

Overall, the exploratory analysis highlights multiple variables that could be useful predictors when building models in the next phase.

# Phase 2: Predictive Modeling

## Step 1: Handle Missing Data
                
```{r}
train_clean <- train %>%
  filter(!is.na(loading)) %>%
  mutate(across(where(is.numeric),
                ~ replace_na(., mean(., na.rm = TRUE))))
```

*Filled in missing numbers with the column mean to keep all the rows*

## Step 2: Train/Test Split

```{r, include=FALSE}
library(tidyverse)
library(tidymodels)
library(caret)
library(mice)
library(VIM)
library(ranger)
library(RColorBrewer)
library(rpart)
library(rattle)
library(e1071)
library(xgboost)
library(usemodels)
library(nnet)
library(stacks)
```

```{r}
set.seed(123)
train_split <- initial_split(train_clean, prop = 0.7, strata = failure)
train_data <- training(train_split)
test_data  <- testing(train_split)
```

## Step 3: Logistic Regression

```{r}
set.seed(123)
folds <- vfold_cv(train_data, v = 5)
```

```{r}
failure_recipe <- recipe(failure ~ ., data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes())
```

```{r}
log_model <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification")

log_workflow <- workflow() %>%
  add_model(log_model) %>%
  add_recipe(failure_recipe)

log_fit <- fit(log_workflow, data = train_data)
```

```{r}
log_preds <- predict(log_fit, test_data, type = "class") %>%
  bind_cols(test_data)

confusionMatrix(log_preds$.pred_class, test_data$failure)
```

## Step 4: Classification Tree

```{r}
tree_recipe <- recipe(failure ~ ., data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes())
```

```{r}
tree_model <- decision_tree(cost_complexity = tune()) %>%
  set_engine("rpart", model = TRUE) %>%
  set_mode("classification")

tree_workflow <- workflow() %>%
  add_model(tree_model) %>%
  add_recipe(failure_recipe)

set.seed(1234)
tree_res <- tree_workflow %>%
  tune_grid(resamples = folds, grid = 25)

saveRDS(tree_res, "tree_res.rds")
```

```{r}
tree_res %>%
  collect_metrics() %>%
  ggplot(aes(cost_complexity, mean)) +
  geom_line(size = 1.5, alpha = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ .metric, scales = "free", nrow = 2)
```

```{r}
final_tree <- finalize_workflow(tree_workflow, select_best(tree_res, metric="accuracy")) %>%
  fit(data = train_data)
```

```{r}
tree_preds <- predict(final_tree, test_data, type = "class") %>%
  bind_cols(test_data)

confusionMatrix(tree_preds$.pred_class, test_data$failure)
```

## Step 5: Random Forest

```{r}
rf_recipe <- recipe(failure ~ ., data = train_data) %>%
  step_dummy(all_nominal(), -all_outcomes())
```

```{r}
rf_model <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 100
) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("classification")

rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

set.seed(1234)
rf_res <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = 30
)

saveRDS(rf_res, "rf_res.rds")
```

```{r}
rf_res %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point() +
  labs(y = "Accuracy", x = "mtry", color = "min_n")
```

```{r}
final_rf <- finalize_workflow(
  rf_workflow,
  select_best(rf_res, metric="accuracy")
) %>%
  fit(data = train_data)
```

```{r}
rf_preds <- predict(final_rf, test_data, type = "class") %>%
  bind_cols(test_data)

confusionMatrix(rf_preds$.pred_class, test_data$failure)
```

*Because failures are relatively rare (21%), model performance was evaluated using sensitivity and specificity in addition to overall accuracy.*

## Step 6: Model Evaluation

```{r}
log_cm  <- confusionMatrix(log_preds$.pred_class, test_data$failure)
tree_cm <- confusionMatrix(tree_preds$.pred_class, test_data$failure)
rf_cm   <- confusionMatrix(rf_preds$.pred_class, test_data$failure)
```

```{r}
model_results <- data.frame(
  Model = c("Logistic Regression", "Classification Tree", "Random Forest"),
  Accuracy = c(log_cm$overall["Accuracy"],
               tree_cm$overall["Accuracy"],
               rf_cm$overall["Accuracy"]),
  Sensitivity = c(log_cm$byClass["Sensitivity"],
                  tree_cm$byClass["Sensitivity"],
                  rf_cm$byClass["Sensitivity"]),
  Specificity = c(log_cm$byClass["Specificity"],
                  tree_cm$byClass["Specificity"],
                  rf_cm$byClass["Specificity"])
)

model_results
```

*The models did a decent job overall, but they weren’t great at identifying the few products that actually failed.*

## Step 7: Kaggle Submission

```{r}
kaggle_test <- read_csv("C:/Users/brewayne/Downloads/test.csv")
```

```{r}
rf_preds_kaggle <- predict(final_rf, new_data = kaggle_test, type = "class")

rf_submission <- data.frame(
  id = kaggle_test$id,
  failure = rf_preds_kaggle$.pred_class
)

write.csv(rf_submission, "rf_submission.csv", row.names = FALSE)
```

```{r}
tree_preds_kaggle <- predict(final_tree, new_data = kaggle_test, type = "class")

tree_submission <- data.frame(
  id = kaggle_test$id,
  failure = tree_preds_kaggle$.pred_class
)

write.csv(tree_submission, "tree_submission.csv", row.names = FALSE)
```